\begin{abstract}

\noindent\textbf{Purpose:}
The apparent diffusion coefficient (ADC) is the de facto biomarker of prostate diffusion MRI, yet the compartment-volume models that explain its success are not measured directly (@Patrick: citations??). We use Bayesian spectral decomposition of extended b-value prostate DWI to quantify which diffusion compartments are reliably recoverable and to explain ADC's near-optimality for region-of-interest (ROI) tumor detection.

\noindent\textbf{Methods:} (@patrick: should we mention the joint inference of noise and spectra to make "fully" clear and excite readers about that (to my knowledge something people have not yet done in spectra reconstruction?? check literature, also should we talk about the sensitivity analysis of ADC that Stephan found cool and novel? lets make sure to not miss advertising cool methodlogical novelties...)
Multi-b DWI ($b = 0$--$3500\;\text{s/mm}^2$, 15 values) from 56 patients (149 ROIs: 40 tumor, 109 normal) was decomposed into eight diffusivity components by constrained-ridge MAP (@Patrick: abbreviation never introduced...non stats people might not know) and fully Bayesian No-U-Turn Sampler (NUTS) inference (@Patrick: is this "sloppy" as NUTS is a sampler while we want to primarily tell the fully bayesian data driven story here? also more in-line with MAP? maybe MAP and NUTS also in fig is not perfect comparison as one is a probabilstic term, the other sampler/method?), the latter providing per-component posterior uncertainty (coefficient of variation, CV) (@patrick: necessary to metnion CV in abstract? just a way to measure ... moer accurate for results / methods or maybe important??). Tumor detection was evaluated through one leave-one-out logistic-regression pipeline (@Patrick: "pipeline" a bit vague? we're talking about creating spectral classifier using logistic regression classification or maybe that's obvious? also we could introduce abbrevaition for leave one out cross validation here or maybe too much detail for abstract? Also what does it mean to apply it to ADC and individual frations, does this even make sense? check with methods section) applied identically to ADC, individual fractions, the two outer fractions ($D = 0.25$ and $3.0\;\umsmm$), and all eight, with patient-level bootstrap confidence intervals. Gleason associations were assessed as grade-dependent spectral shifts ($n = 29$ graded).

\noindent\textbf{Results:}
Only restricted diffusion (CV = 0.20) and free water (CV = 0.32) were well identified; intermediate components had CV $> 0.80$, a data limit robust to prior width. The spectral classifier was statistically indistinguishable from ADC, and a two-fraction model matched the full eight (NUTS AUC 0.93--0.94; ADC raw rank 0.95--0.98): detection signal collapses onto the two outer compartments. ADC anti-correlated with the spectral discriminant at $|r| > 0.95$ for both estimators. Grade-dependent shifts dissociated a detection axis (free water collapsing at tumor onset) from a grading axis (restricted and glandular fractions tracking grade) (@patrick: not sure this statement is not too bold for what we see, should we actually talk about "two axes" here, since middle compartments did not show any notable contribution for detection...I do think that there is some prior work here that we coudl reference...ask Stephan and check latest discussion section).

\noindent\textbf{Conclusion:}
At the ROI level the diffusivity spectrum is diagnostically equivalent to ADC, which explains why this scalar is so hard to improve upon (@Patrick: I find this too broad for the first sentecne of the conclusion, this should be the "sensational" news or not? the major finding is the spectral classifier collapses to two bins that can be uniquely described by ADC as they move in one motion (.25 increases, while 3.0 decreases..., also I just say "so hard to improve upon" shouldn't this be the crucial part where I either disprove to a certain degree pepole claiming to outperform ADC or at least cite people openly struggeling with improving upon it?). Bayesian decomposition does not obsolete ADC but mechanistically explains it while quantifying compartment recoverability and per-component uncertainty; tuned MAP and NUTS yield equivalent spectra, so the Bayesian gain is calibrated uncertainty, not better point estimates (@patrick: important statement but frames it somewhat in a negative argumentative way? just say map is fast computation while nuts epxloratory tool with free unceratinty estimates...?).

\end{abstract}
