\section{Discussion}
The central finding is that ADC, despite collapsing the multi-component signal to a single scalar, is a near-optimal spectral discriminant for tumor detection: across zones and estimators it ranks ROIs almost identically to the optimal eight-component classifier, because both concentrate on the two well-identified outer compartments (Figures~\ref{fig:adc_discriminant} and~\ref{fig:sensitivity}).

Rather than an impoverished summary of a richer signal, ADC effectively captures the dominant axis of spectral variation relevant to tumor detection.
The spectral decomposition does not obsolete ADC; it explains it.
The added value of multi-component analysis lies in three areas: biological interpretability (which compartments---restricted diffusion, free water---drive the ADC contrast); uncertainty quantification (which components are reliably estimated); and the potential to surface grade-dependent spectral shifts that ADC compresses, such as glandular-water differences between Gleason grades.

\stephan{candidate interpretation, not a validated finding --- please confirm it is histologically sound.} A candidate histological interpretation of the grade-dependent spectral shifts (Figure~\ref{fig:spectrum_ggg}) follows from the cellular reorganization characteristic of prostate adenocarcinoma.
The dominant free-diffusion bin at $D = 3.0\;\umsmm$---bulk water in gland lacunae of $\approx$100--300\,\textmu m---collapses in both tumor groups, consistent with replacement of patent lumina by solid tissue.
The $D = 2.0\;\umsmm$ bin is more grade-specific: low-grade GGG\,$=\,1$ lesions retain near-normal glandular architecture and $D = 2.0\;\umsmm$ mass, whereas the loss of glandular structure at higher Gleason patterns collapses this compartment too.
The redistribution of displaced spectral mass toward $D = 0.25\;\umsmm$, most pronounced in GGG\,$\geq 3$, is consistent with the increased cell-packing density that is the morphological hallmark of higher-grade cancer, and appears to track the predominant-Gleason-pattern transition at which ADC-based grading is itself least reliable---although the small per-grade samples preclude a statistical claim.
Any incremental grading signal beyond ADC would have to come from the intermediate bins, where ADC is least sensitive (Figure~\ref{fig:sensitivity}); but their wide posterior uncertainty (CV~$> 0.8$) and the small grade-specific samples preclude a statistical claim.
We present this histology mapping as a candidate interpretation the decomposition makes directly observable, not a validated finding; testing it would require a larger cohort with whole-mount histology correlation.

The comparison between MAP and NUTS estimation revealed a complementary relationship.
At tuned regularisation ($\lambda = 10^{-3}$), MAP and NUTS agreed closely for the well-identified components and their discriminant scores were nearly interchangeable ($r > 0.98$), confirming that closed-form MAP is sufficient for classification and recovers approximately the same fractions as the full sampler on realistic spectra.
The divergence between MAP and NUTS on intermediate components ($D = 0.5$--$1.0\;\umsmm$) was not a deficiency of either method but a faithful reflection of the ill-posedness: these components are genuinely undetermined by 15 b-value measurements, even at the much lower noise levels achieved with the endorectal coil and low-resolution acquisition of the present cohort.
NUTS makes this ill-posedness explicit through wide posterior intervals (CV $> 0.80$), reflecting the same Bayesian gain (effective estimation-error reduction relative to the unconstrained Cram\'er-Rao bound) that makes the well-identified components quantitatively trustworthy; MAP, in contrast, silently returns a point estimate that may be misleadingly precise.
The Bayesian approach thus identifies what can and cannot be learned from a given acquisition \citep{prange2009quantifying}.
Propagating the posterior through the classifier did not add diagnostic value beyond the point estimate (Figure~\ref{fig:uncertainty}): consistent with the central finding, detection rides on the two well-identified outer compartments, leaving little estimation uncertainty in the discriminant direction to propagate. The posterior's value is therefore upstream---establishing which compartments are reliably recovered---not in the tumor-versus-normal decision itself.

The degeneracy of the intermediate components is a fundamental limit of the acquisition, not an artifact of the inference or the discretization: the Fisher correlations are intrinsic to the b-value sampling, the sampler reproduces the same smearing on simulated ground truths (Figure~\ref{fig:validation}), and re-fitting on grids that honor the $\Delta D \propto D^{3/2}$ spacing does not recover the lost components. What makes the well-constrained components quantitatively trustworthy is the two-step tightening of the achievable estimation error in Figure~\ref{fig:fisher}b, which we read as two related gains: the prior contributes a regularization gain (the unconstrained-to-van-Trees step), and non-negativity contributes a further constraint gain (the van-Trees-to-NUTS step). Both act in the same direction---pulling the estimate toward the feasible, physically meaningful region---and together they reduce the effective estimation error by orders of magnitude relative to the unconstrained Cram\'er--Rao bound. Multi-exponential diffusion separation is itself ill-conditioned, with Cram\'er--Rao analyses showing severe noise propagation when the decaying components are close \citep{jalnefjord2019optimization}; the constrained Bayesian estimator sidesteps this by working in a far better-posed regime, reconciling that long-recognized difficulty with the recoverability we observe.

This study has several limitations.
First, the tumor ROIs were placed based on multiparametric MRI (including DWI/ADC maps), introducing a potential circularity: lesions that are well delineated on ADC maps may be overrepresented \citep{langkilde2018evaluation}.
This bias could inflate ADC's apparent discriminative performance relative to spectral features.
Only the Gleason grade classification, derived from biopsy pathology, is free from this circularity.
Second, the GGG sample size ($n = 29$) is insufficient for reliable AUC comparison between methods; differences below $\Delta\text{AUC} \approx 0.10$ cannot be meaningfully distinguished.
Third, the spectral model assumes a fixed set of diffusivity bins; a continuous spectral approach \citep{prange2009quantifying} might reveal additional structure but at the cost of increased ill-posedness.
Fourth, only one fixed echo-time/diffusion-time configuration was analyzed; the shorter echo and diffusion times enabled by recent gradient hardware \citep{zhu2024human, elsaid2026nonlinear} could change both the SNR and the relative weight of the compartments visible to the decomposition.
Finally, the pixel-wise demonstration (Figure~\ref{fig:pixelwise}) was limited to a single supplementary patient; generalization requires larger studies with histopathological correlation.
It also exposes a calibration caveat for per-voxel deployment: applying the ROI-trained detector voxel-wise yields scores that skew systematically toward the tumor class, because the substantially lower SNR of a single voxel relative to an averaged ROI inflates the restricted-diffusion ($D = 0.25\;\umsmm$) fraction.
This reinforces our decision to report all quantitative results at the ROI level, and indicates that faithful per-voxel spectral classification would require either explicit voxel-level noise modeling or spatial regularization.

The framework presented here is directly applicable to other organs and tumor types where multi-b DWI is acquired, including breast, liver, and brain tumors.
Amortized inference---training a neural network to approximate the posterior---could reduce computation from minutes per ROI to milliseconds, enabling real-time deployment, and the Fisher information framework could guide joint optimization of b-value protocols and discretization grids for specific clinical questions.
Integration with T\textsubscript{2}-weighted imaging could yield multi-parametric spectral biomarkers.

% =====================================================================
% STAGING AREA --- PARKED CONTENT FOR THE DISCUSSION REWORK (2026-06-23)
% NOT RENDERED (\iffalse...\fi). Verbatim prose moved here during the Results
% pass: (a) the sensitivity-mechanism + GGG + prediction-confidence + pixel-wise
% subsections cut from Results, and (b) the ADC-Sensitivity + Pixel/Direction
% subsections cut from Methods. To be reshaped/merged into the Discussion
% paragraphs above (which already cover much of this), then deleted.
% =====================================================================
\iffalse

%% ---- (a1) FROM RESULTS: mechanism half of the old "Why ADC works" ----
The mechanism is visible in the per-bin structure of the discriminant (Figure~\ref{fig:sensitivity}).
The ADC sensitivity vector $\partial\text{ADC}/\partial R_j$ at the average tumor spectrum is strongly negative for restricted diffusion ($-1.59$: a larger restricted fraction lowers ADC) and positive for free water ($+0.85$); because the logistic-regression coefficients point the opposite way, Figure~\ref{fig:sensitivity} plots the sign-flipped $-\partial\text{ADC}/\partial R_j$ alongside them, so aligned bars indicate matching direction.
Using NUTS features, the two vectors align well (Pearson $r = -0.79$, PZ, $95\%$ CI $[-0.89,\,-0.50]$; $r = -0.88$, TZ, $[-0.95,\,-0.70]$), and the alignment is carried by the two outer bins: only $D = 0.25$ and $D = 3.0\;\umsmm$ have logistic coefficients whose bootstrap intervals exclude zero.
This is not because ADC ignores the intermediate compartments: its sensitivity $-\partial\text{ADC}/\partial R_j$ is appreciable there (e.g.\ $D = 0.5\;\umsmm$), since ADC is a smooth, monotonic-in-$D$ weighting of the entire spectrum. The resolution is that discriminative power is sensitivity \emph{times} tumor--normal contrast: the intermediate fractions carry little contrast beyond what the outer bins already capture (Table~\ref{tab:auc}) and are poorly identified (high CV), so ADC's sensitivity to them is spent on a low-contrast, noisy direction that shifts neither score's ranking. Both summaries therefore reduce to the same two outer compartments, which is why the per-ROI scores rank almost identically (Figure~\ref{fig:adc_discriminant}) even though the per-bin profiles diverge in the middle. The divergence is confined to bins that are diagnostically inert for both---it is not evidence that ADC and the spectral classifier estimate different quantities.

%% ---- (a2) FROM RESULTS: Spectral Decomposition by Gleason Grade Group ----
Figure~\ref{fig:spectrum_ggg} shows the cohort's average NUTS posterior-mean spectrum by Gleason Grade Group, with normal tissue ($n = 109$) as the shared baseline and tumor resolved into GGG\,$=\,1$ ($n = 8$), GGG\,$=\,2$ ($n = 12$), and GGG\,$\geq 3$ ($n = 9$).
\stephan{exploratory subsection --- please confirm the intermediate-bin interpretation below; the independent grading-axis claim is dropped for now.} These grade-stratified spectra are exploratory: the small, unevenly graded subgroups (down to $n = 8$) do not support a grading axis independent of detection, but the per-bin trends below are suggestive.
The free-diffusion bin at $D = 3.0\;\umsmm$ (intact gland lumen) is suppressed as soon as tumor is present ($0.39$ normal $\to 0.21$ at GGG\,$=\,1$) and then changes little ($0.16$ at GGG\,$\geq 3$): it saturates at tumor onset, separating tumor from normal but not grade from grade.
By contrast the restricted-cellular bin ($D = 0.25\;\umsmm$) rises with grade ($0.07$ normal to $0.31$ at GGG\,$\geq 3$) and the glandular-epithelial bin ($D = 2.0\;\umsmm$) is preserved in low-grade tumor then depleted ($0.26$ normal, $0.28$ at GGG\,$=\,1$, $0.17$ at GGG\,$\geq 3$): these bins, where ADC is least sensitive (Figure~\ref{fig:sensitivity}), track aggressiveness.
Grade-specific sample sizes are small (down to $n = 8$--$9$) and the corresponding confidence bands wide, and 11 of the 40 tumor ROIs (7 with biopsy Gleason Grade Group 0 and 4 ungraded) lack a Grade Group and are excluded from this grade stratification, though they remain tumor ROIs in the detection analysis.
Method robustness: re-running the same group averaging on tuned-MAP ($\lambda = 10^{-3}$) posterior estimates rather than NUTS posterior means yielded indistinguishable group trajectories on all eight bins.

%% ---- (a3) FROM RESULTS: Prediction confidence ----
Each classifier returns a probability $P(\mathrm{tumor})$ rather than a hard label, and misclassified ROIs concentrate near the decision boundary ($P \approx 0.5$): ranking predictions by their distance from the boundary separates likely errors from confident, correct calls, so the classifier flags its own least-reliable predictions (Figure~\ref{fig:uncertainty}).
This confidence is a property of the logistic model, available from the point estimate alone, and is shared by the ADC- and spectrum-based classifiers.
Propagating the full NUTS posterior through the classifier---applying it to every posterior draw rather than the posterior mean---does not add to this.
The 90\% credible interval on $P(\mathrm{tumor})$ is on average $2.4\times$ wider for misclassified than for correctly classified ROIs (pooled across zones, mean width $0.39$ vs $0.16$), but this reflects the logistic geometry rather than the spectral posterior: the slope $P(1-P)$ is largest at the boundary, so in logit space the difference falls to $\approx 1.3\times$ and is no longer significant, and the interval width does not predict misclassification once distance to the boundary is accounted for.
Offering the per-component posterior standard deviations to the classifier as features likewise leaves detection AUC essentially unchanged.
The posterior therefore quantifies which spectral components are reliably estimated (identifiability), but adds no diagnostic information to the tumor-versus-normal decision beyond the point estimate.

%% ---- (a4) FROM RESULTS: Pixel-wise Spectral Decomposition ----
Pixel-wise spectral estimation was performed on the manually segmented prostate of the supplementary patient (MAP and NUTS; all NUTS voxels converged, $\hat{R} < 1.05$).
At the voxel level the restricted-diffusion ($D = 0.25\;\umsmm$) and free-water ($D = 3.0\;\umsmm$) fraction maps were spatially anti-correlated (Figure~\ref{fig:pixelwise}), and the per-voxel spectral discriminant (ROI-level PZ coefficients applied to the pixel spectra) reproduced the ADC pattern (anti-correlation $r = -0.97$).
The contribution unique to the Bayesian decomposition is the accompanying per-voxel uncertainty map: propagating the NUTS posterior through the classifier flags where a voxel's score is least reliable, and the eight-bin detector is on average $1.8\times$ more uncertain than the two-bin detector because it weights the unidentifiable intermediate compartments---an uncertainty that ADC and point-estimate methods cannot provide.
As at the ROI level, the discriminant adds no detection signal beyond ADC; its value here is the calibrated, spatially-resolved confidence it carries.
Direction independence was verified at the whole-ROI level on a separate three-direction acquisition (Figure~\ref{fig:directions}): across all 13 regions the per-direction coefficient of variation of the estimated spectrum was lowest for the diagnostically dominant restricted-diffusion bin ($D = 0.25\;\umsmm$, mean $9\%$) and was large only on the near-empty bins ($D = 20\;\umsmm$), where it reflects a small-denominator effect rather than genuine anisotropy, confirming that trace averaging is appropriate for spectral decomposition.

%% ---- (b1) FROM METHODS: ADC Sensitivity Analysis ----
The ADC sensitivity vector (Eq.~\ref{eq:sensitivity}) was computed numerically at the average tumor and average normal spectra for each zone, using finite differences with step size $\epsilon = 10^{-4}$.
For the sensitivity-alignment comparison, the logistic-regression \emph{coefficient vector} was taken from a single full-data fit of the eight-feature model---distinct from the leave-one-out detection pipeline above, which evaluates predictive AUC---since only the coefficient direction, not predictive performance, enters this analysis.
The Pearson and Spearman rank correlations between the sensitivity vector and the LR coefficient vector were computed to quantify alignment.
Bootstrap 95\% confidence intervals for ROI-level correlations between ADC and the spectral discriminant were obtained by resampling ROI pairs with replacement (1000 resamples).

%% ---- (b2) FROM METHODS: Pixel-wise and Direction-wise Spectral Estimation ----
To demonstrate the framework at the voxel level, pixel-wise decomposition was performed on a single axial slice from one patient, with the prostate manually segmented. Each voxel's decay, normalized by its $b = 0$ signal, was decomposed by MAP (Eq.~\ref{eq:map}) and NUTS (Eqs.~\ref{eq:prior_R}--\ref{eq:likelihood}) with the ROI-level parameters, and a spectral discriminant score was computed by applying the peripheral-zone classifier coefficients to the pixel spectra (the segmented tumor lay in the peripheral zone); for NUTS, discriminant uncertainty was propagated by applying the classifier to each of 500 posterior draws per voxel, yielding a per-voxel distribution of scores.
To verify the trace-averaging assumption used at the ROI level, a separate acquisition that retained the per-encoding-direction signal---13 ROIs across 7 patients, with the same 15 b-values along three orthogonal directions---was decomposed independently for each direction with the MAP estimator and compared to the trace-averaged spectrum. The aggregate per-direction coefficient-of-variation statistics reported in Results are computed across all 13 ROIs; Figure~\ref{fig:directions} shows a representative subset of four patients (one peripheral-zone and one transition-zone ROI each).

\fi
% ===================== END STAGING AREA =====================
