% === FIGURE LEGENDS ===
% Listed separately from the figure floats (MRM style: captions go in a list at
% the end of the manuscript text, not inside the figure files). Each number is
% auto-synced to its float via \ref to the figure's \label, so the list stays
% correct even if the figures are reordered.
%TC:ignore
\section*{Figure Legends}

\noindent\textbf{Figure~\ref{fig:fisher}.}
Fisher-information analysis of the eight-bin spectral discretization.
(a)~Fisher-information correlation matrix
$C_{jk} = F_{jk}/\sqrt{F_{jj}F_{kk}}$: adjacent intermediate components
($D = 0.50$--$1.00\;\umsmm$) are mutually correlated above $0.99$, making them
nearly indistinguishable from the 15 b-value measurements, whereas
$D = 0.25$ and $D = 20\;\umsmm$ are the most separable.
(b)~Per-component estimation uncertainty at the cohort-median SNR $= 303$: the
unconstrained Cram\'er--Rao lower bound (light gray), the Bayesian (van Trees)
CRLB with the half-normal prior (dark gray), and the empirical NUTS posterior
standard deviation (orange).
Two improvement factors are annotated on the bars---the prior gain (gray,
unconstrained\,$\to$\,Bayesian) and the constraint gain (orange,
Bayesian\,$\to$\,NUTS, up to $74\times$)---which together reduce the estimation
uncertainty by up to three orders of magnitude relative to the unconstrained
bound, because the prior and the non-negativity constraint truncate the
feasible region.
(c)~Individual component decay curves $R_j \exp(-bD_j)/K$ at equal fractions,
with horizontal noise floors at SNR $= 50$, $100$, and $303$: slow components
($D = 0.25\;\umsmm$) contribute signal across all b-values, whereas fast
components ($D \geq 3.0\;\umsmm$) fall below the noise floor within the first
few b-value steps, motivating the sparser discretization at high diffusivities.
\sandy{panel (b) van-Trees Bayesian-CRLB pending your derivation validation.}

\medskip\noindent\textbf{Figure~\ref{fig:spectra}.}
Cohort diffusivity spectra by tissue type and anatomical zone.
For each diffusivity bin, box plots show the cohort distribution of the
estimated water fraction $R_j$ across ROIs (box: interquartile range; line:
median; whiskers: $1.5\times$ IQR), with MAP (green) alongside NUTS
posterior-mean estimates (orange) at the tuned regularization
$\lambda = 10^{-3}$.
Top row: normal; bottom row: tumor. Left column: peripheral zone (PZ); right:
transition zone (TZ).
Relative to normal tissue, tumor shows an elevated restricted-diffusion
fraction ($D = 0.25\;\umsmm$) and a reduced free-water fraction
($D = 3.0\;\umsmm$) in both zones, and MAP and NUTS agree closely.
Each box is the cohort distribution of the per-ROI posterior-mean fraction
(between-ROI spread); within-ROI posterior uncertainty is shown in
Figure~\ref{fig:sensitivity} and the supplementary atlas.

\medskip\noindent\textbf{Figure~\ref{fig:validation}.}
Validation of the spectral inference on simulated ground truth across spectrum
shape and noise level.
Six ground-truth spectra (rows A--F: normal-like and tumor-like---the
cohort-mean NUTS spectra---a uniform spectrum, a concentrated $\delta$ at
$D = 0.75\;\umsmm$, an inverse-tumor (reversed tumor), and a bimodal spectrum)
are each recovered at SNR $= 50$, $300$, and $1000$ (columns); SNR $300$ is the
cohort median (the operating point of the real data) and $50$ reflects the
reduced per-voxel SNR of the pixel-wise analysis.
In every panel the bars are the ground truth (grey), the tuned-MAP estimate
($\lambda = 10^{-3}$, green), and the NUTS posterior mean (orange); error bars
% TODO: set n to the final rep count once the 100-rep run completes
are $\pm 1$ standard deviation across $n = 100$ independent noise realizations,
each fit with the deployed NUTS configuration (the per-panel
$\hat{R}_{\max}$ is the worst split-$\hat{R}$ over those realizations).
MAP and NUTS recover the smooth spectra equivalently and tighten with SNR,
whereas the $\delta$ spike is only partially recovered and smeared into its
correlated neighboring bins---the identifiability limit of this b-value grid,
not a sampler artifact; on that concentrated spectrum NUTS recovers more of the
peak than MAP.
The bottom row reports noise recovery for each ground truth (A--F): the true
$\sigma = 1/\mathrm{SNR}$ (dashed) against the MAP residual estimate and the
NUTS-inferred $\hat{\sigma}$, both close to truth---the noise level is
recoverable with or without a sampler.

\medskip\noindent\textbf{Figure~\ref{fig:roc}.}
Tumor detection collapses onto the two outer diffusivity compartments, and the
spectral classifier is equivalent to ADC.
Leave-one-out cross-validated ROC for tumor-versus-normal detection in the
peripheral zone (PZ, left; $n = 81$) and transition zone (TZ, right;
$n = 68$). The three thick reference curves---ADC (black), the spectral
classifier on all eight fractions (orange solid) and on only the two outer
fractions $\{D = 0.25,\,3.0\;\umsmm\}$ (orange dashed)---pass through the
identical NUTS-feature leave-one-out logistic-regression pipeline ($C = 1.0$)
for an apples-to-apples comparison.
The thin single-component curves are each a raw single-feature ROC in the
tumor-positive orientation (not the LOOCV pipeline): the strong outer
components $D = 0.25\;\umsmm$ (restricted, teal) and $D = 3.0\;\umsmm$
(free water, brown); three weak single bins whose weakness is zone-dependent,
$D = 2.0\;\umsmm$ (goldenrod, near chance in PZ), $D = 1.5\;\umsmm$ (indigo,
near chance in TZ), and $D = 20\;\umsmm$ (magenta, near chance in both); and
the remaining intermediate bins as a faint grey bundle.
The three thick curves are statistically indistinguishable and the
two-fraction classifier matches the full eight-fraction one: the intermediate
and free-water fractions add nothing to detection.
AUC values with 95\% confidence intervals are reported in
Table~\ref{tab:auc}.

\medskip\noindent\textbf{Figure~\ref{fig:adc_discriminant}.}
ADC is a near-optimal spectral discriminant, and the equivalence holds for
both estimators.
ADC versus the spectral discriminant score (logistic-regression linear
predictor from the eight spectral fractions) across ROIs, by anatomical zone
(columns: PZ, $n = 81$, left; TZ, $n = 68$, right) and spectral estimator
(rows: NUTS posterior-mean, top; constrained-MAP at tuned
$\lambda = 10^{-3}$, bottom).
All panels share common axes; the discriminant is fit separately per zone and
estimator, so its score scale is comparable only within a panel, whereas ADC
($x$-axis) is the same physical quantity throughout.
Tumor ROIs (red) cluster at low ADC and high discriminant, normal ROIs (blue)
the reverse; the black line is the ordinary-least-squares (OLS) fit.
Per-panel Pearson $r$ with a 95\% percentile-bootstrap confidence interval is
annotated in each panel (all $|r| > 0.95$).
The near-perfect anti-correlation shows that ADC rank-orders ROIs almost
identically to the spectral classifier, independent of estimator.

\medskip\noindent\textbf{Figure~\ref{fig:sensitivity}.}
The spectral discriminant is statistically a two-compartment detector, and its
two stable bins coincide with where ADC is most sensitive.
For each zone (PZ, TZ), bars give the standardized tumor-versus-normal
logistic-regression coefficient per diffusivity bin (NUTS features, $C = 1.0$,
the same fit as Figure~\ref{fig:adc_discriminant}), normalized to
$\max|w| = 1$, with 95\% bootstrap confidence intervals.
Charcoal diamonds give the ADC sensitivity $-\partial\text{ADC}/\partial R_j$
at the average tumor spectrum (normalized, sign-aligned to the tumor
direction).
Each bar is colored and hatched by the bin's within-ROI posterior CV (light
$=$ well identified to dark $=$ poorly identified), a per-ROI identifiability
distinct from the cohort-level coefficient stability shown by the bootstrap
intervals.
In both zones only the two outer bins carry stable weight ($D = 0.25\;\umsmm$
positive, $D = 3.0\;\umsmm$ negative); the intermediate coefficients are
individually unstable from inter-bin collinearity and poor identifiability, so
their separate detection weights are not attributable.
The profile anti-correlates with $\partial\text{ADC}/\partial R_j$
($r = -0.79$ PZ, $-0.88$ TZ), carried by the two outer bins; the moderate
whole-profile value reflects the noisy intermediate coefficients, which carry
little weight either way.

\medskip\noindent\textbf{Figure~\ref{fig:spectrum_ggg}.}
The diffusivity spectrum shifts systematically across Gleason Grade Groups.
NUTS posterior-mean spectrum averaged across ROIs, in two contrasts that share
a normal-tissue baseline (gray, $n = 109$). Left (emergence): GGG\,$=\,1$
($n = 8$) versus GGG\,$\geq 2$ ($n = 21$). Right (aggressiveness): GGG\,$=\,2$
($n = 12$) versus GGG\,$\geq 3$ ($n = 9$). Each grade group has its own color.
Shaded bands are 95\% confidence intervals of the group mean (Student's $t$,
$\text{df} = n - 1$).
The free-water bin ($D = 3.0\;\umsmm$, a proxy for intact gland lumen)
collapses at tumor onset ($0.39$ normal $\to 0.21$ at GGG\,$=\,1$) and changes
little thereafter ($0.16$ at GGG\,$\geq 3$): it carries detection, not grade.
In contrast the restricted-cellular bin ($D = 0.25\;\umsmm$) and the
glandular-epithelial bin ($D = 2.0\;\umsmm$), where ADC is least sensitive,
track grade: $D = 0.25$ rises with grade ($0.07$ normal, $0.17$ at
GGG\,$=\,1$--$2$, $0.31$ at GGG\,$\geq 3$) and $D = 2.0$ is preserved in
low-grade tumor then depleted ($0.26$ normal, $0.28$ at GGG\,$=\,1$, $0.17$ at
GGG\,$\geq 3$).
Grade groups are small (down to $n = 8$--$9$) and the bands wide; 11 of 40
tumor ROIs (7 with GGG\,$=\,0$, 4 ungraded) are not shown.
\stephan{exploratory --- please confirm the grade-dependent reading of the
intermediate bins ($D = 0.25$, $2.0\;\umsmm$); the independent grading-axis
claim has been dropped pending a larger graded cohort.}

\medskip\noindent\textbf{Figure~\ref{fig:uncertainty}.}
The classifier returns a calibrated probability of cancer per ROI, and its
confidence---a property of the classifier, not of the spectral posterior---flags
the predictions most likely to be wrong.
For each ROI the leave-one-out classifier (the same NUTS-feature pipeline as
Figure~\ref{fig:roc}, $C = 1$) is applied to every NUTS posterior draw of the
eight fractions, yielding a posterior distribution over $P(\mathrm{tumor})$.
Per-ROI predictions in the peripheral zone (top, $n = 81$) and transition zone
(bottom, $n = 68$), each sorted by predicted probability: markers at the
posterior-mean probability (circles normal, squares tumor) with error bars the
propagated 90\% credible interval; correctly classified ROIs in pale colors,
misclassified in saturated colors.
Misclassified ROIs cluster near the decision boundary ($P = 0.5$), where the
interval is on average $2.4\times$ wider (pooled mean width $0.39$ vs $0.16$).
This widening is a geometric property of the logistic link---its slope
$P(1-P)$ is largest at the boundary---not a contribution of the spectral
posterior: in logit space the misclassified-to-correct difference is not
significant, and supplying the per-component posterior standard deviations to
the classifier as features leaves detection AUC unchanged.
The confidently-misclassified ROIs are almost all low-grade or ungraded tumors
whose spectra resemble normal tissue (cf.\ Figure~\ref{fig:spectrum_ggg})---a
limit of diffusion contrast, not a miscalibration.
The classifier's per-prediction confidence is thus a useful triage signal
available already from the point estimate.

\medskip\noindent\textbf{Figure~\ref{fig:pixelwise}.}
Pixel-wise spectral decomposition of a single peripheral-zone prostate DWI
slice (anonymized patient), showing that the ROI-level detector extends to the
voxel level while uniquely providing a per-voxel uncertainty map.
(A)~Anatomical reference ($b = 0$).
(B,\,C)~NUTS posterior-mean restricted-diffusion ($D = 0.25\;\umsmm$) and
free-water ($D = 3.0\;\umsmm$) fractions.
(D)~Conventional ADC map (restricted diffusion appears dark).
(E,\,F)~Per-voxel discriminant score for the two-bin
$\{D = 0.25,\,3.0\;\umsmm\}$ and full eight-bin classifiers (peripheral-zone
detectors, same recipe as Figure~\ref{fig:roc}), centered on the decision
boundary; both reproduce the ADC pattern ($r = -0.93$ and $-0.96$). The
absolute score skews tumor-like because the lower SNR of a single voxel
inflates the restricted fraction, which is why quantitative results are
reported at the ROI level.
(H,\,I)~Posterior standard deviation of the score (500 Monte Carlo draws) for
the two- and eight-bin detectors on a shared scale: the eight-bin map is
$1.8\times$ more uncertain because it weights the unidentifiable intermediate
bins---an uncertainty ADC and point-estimate methods cannot provide.
%TC:endignore
