\section{Methods}

\subsection{Patient Cohort and MRI Acquisition}

Retrospective multi-b DWI data were obtained from the cohort previously described by \citet{langkilde2018evaluation}, acquired under an institutional review board--approved protocol; all analyses used region-of-interest data containing no patient-identifying information.
For the 56 included patients, age ranged from 43 to 77 years (mean 62 years) and the mean serum prostate-specific antigen (PSA) level was 8.8~ng/mL (range, 0.37--46~ng/mL).
The cohort spanned the diagnostic spectrum encountered at biopsy, including confirmed adenocarcinoma as well as benign and high-grade prostatic intraepithelial neoplasia (HGPIN) findings.

All patients underwent a standard multiparametric MR examination with an additional diffusion-weighted sequence on a GE 3~T Discovery MR750 scanner (GE Medical Systems, Milwaukee, WI).
An endorectal coil (Medrad, Pittsburgh, PA) was used in combination with pelvic phased-array coils for signal reception, providing high signal-to-noise ratio at the cost of a non-routine setup.
The added DWI sequence used a matrix size of $64 \times 64$, field of view $280 \times 280\;\text{mm}^2$, slice thickness 5~mm without gap, TR~4000~ms, and TE~100~ms.
The relatively long echo time was required to accommodate the high diffusion weighting with gradient times $\delta = 37$~ms and $\Delta = 45$~ms.
Fifteen b-values were acquired in 250~s/mm$^2$ increments from $b = 0$ to $3500\;\text{s/mm}^2$ along each of three orthogonal diffusion-encoding directions, combined into a trace decay (geometric mean across directions) per b-value; total scan time for the extended-range acquisition was 3~min.

On the multiparametric images an expert radiologist delineated, for each patient, a focal suspicious lesion (the mpMRI target) where present, together with reference normal-appearing tissue in the peripheral and transition zones; these segmentations defined the regions of interest (ROIs).
Within each ROI the trace signal was averaged across the segmented voxels at each b-value to yield the ROI signal-decay curve, which was paired with the patient's biopsy Gleason score.
A total of 149 ROIs were analyzed: 40 mpMRI-defined targets and 109 normal-tissue ROIs, spanning the peripheral zone (PZ; 81 ROIs: 27 target, 54 normal) and transition zone (TZ; 68 ROIs: 13 target, 55 normal).
The targets form the positive class for tumor detection. Of the 40, 29 were biopsy-confirmed adenocarcinoma with an assigned Gleason Grade Group (GGG); the remaining 11 comprised 7 benign, 1 HGPIN, and 3 without a separately recorded Gleason score.
Detection thus distinguishes radiologically suspicious lesions from normal tissue rather than histologically confirmed cancer alone, a point we return to in the Discussion.
Biopsy GGG labels were available for the 29 confirmed-cancer targets (20 low-grade, GGG~1--2; 9 high-grade, GGG~3--5; PZ 21, TZ 8), which form the cohort for the grading analysis.
A per-ROI SNR was derived from the voxel count $N$ as $\mathrm{SNR} = 150\sqrt{N/16}$ and used only as a descriptive index of ROI reliability (cohort median 303, interquartile range 176--478) and as the representative operating point for the Fisher-information analysis (Figure~\ref{fig:fisher}); it is not a fixed noise level in estimation---NUTS infers the per-ROI noise jointly with the spectrum, and the MAP estimate depends only on the ridge weight (Theory).

\subsection{Spectral Estimation}

For each ROI, the trace signal decay was decomposed into eight diffusivity components ($D \in \{0.25, 0.5, 0.75, 1.0, 1.5, 2.0, 3.0, 20.0\}\;\umsmm$) using both MAP and fully Bayesian estimation as described in Theory.

MAP spectra were computed via Eq.~\ref{eq:map} and normalized to sum to one (fractional representation); the constrained least-squares solve used SciPy \citep{virtanen2020scipy} and was instantaneous ($< 1\;\text{s}$ for all 149 ROIs).
The ridge weight $\lambda$ governs a bias--variance trade-off that materially shapes the recovered spectrum: too small a value lets the ill-conditioned fit chase noise, whereas too large a value over-smooths, pulling mass out of the well-determined outer bins and spreading it across the correlated intermediate bins. We set $\lambda = 10^{-3}$ by maximizing recovery of simulated ground-truth spectra (below).

The Bayesian posterior (Eqs.~\ref{eq:prior_R}--\ref{eq:likelihood}) was sampled independently for each of the 149 ROIs using NUTS via PyMC \citep{abrilpla2023pymc} under the half-normal spectrum prior (scale $\sigma_R = 3.16$, precision $1/\sigma_R^2 = 0.1$) and a weakly informative half-Cauchy noise prior (scale $0.01$), with 2000 post-tuning draws, 200 tuning steps, 4 independent chains, and target acceptance rate 0.95 (above the PyMC default of 0.8, to limit divergences on the correlated posterior).
The noise level $\sigma$ was inferred jointly with the spectrum for each ROI rather than fixed from the acquisition.
Convergence was assessed using the rank-normalized split-$\hat{R}$ statistic \citep{vehtari2021rank}, computed with ArviZ \citep{kumar2019arviz}; all 149 ROIs converged ($\hat{R} < 1.05$).
Posterior means of the spectral fractions (normalized to sum to one) were used as NUTS point estimates; posterior standard deviations quantify per-component uncertainty, summarized per component by the cohort-averaged coefficient of variation (CV $=$ posterior std / posterior mean).

Both estimators were validated against simulated data spanning spectrum shape and noise level (Figure~\ref{fig:validation}): six ground-truth spectra---the cohort-mean normal and tumor spectra, a uniform spectrum, a concentrated $\delta$ spike, a reversed-tumor spectrum, and a bimodal spectrum---were each reconstructed at SNR 50, 300 (the cohort median), and 1000 over 100 noise realizations.
On the smooth, physiologically realistic spectra, MAP at $\lambda = 10^{-3}$ and NUTS recovered the ground truth equivalently and tightened with increasing SNR; the concentrated $\delta$ spike was only partially recovered and smeared into its correlated neighboring bins---the identifiability limit of this b-value grid (Theory), not a sampler artifact---with NUTS recovering more of the peak than MAP.
The jointly inferred NUTS noise level tracked the truth ($\hat{\sigma}/\sigma_{\mathrm{true}} \approx 1$) across all shapes, whereas the post-hoc MAP residual under- or over-estimated $\sigma$ depending on spectrum shape, confirming that the noise level is better quantified by the joint Bayesian inference than by a residual plug-in.

\subsection{ADC Computation}

Apparent diffusion coefficients were computed by monoexponential least-squares fitting of $\log S(b) = \log S_0 - b \cdot \text{ADC}$ to the raw signal at b-values $\leq 1000\;\text{s/mm}^2$ (five measurements per ROI: $b = 0, 250, 500, 750, 1000\;\text{s/mm}^2$).
This follows PI-RADS~v2.1, which specifies a maximum b-value of 800--1000~s/mm$^2$ for ADC maps \citep{turkbey2019prostate, weinreb2016pirads}; the b-value range materially affects the absolute ADC value \citep{maier2022choice}.

\subsection{Classification}

Tumor detection was evaluated by zone with $L_2$-regularized logistic regression under leave-one-out cross-validation (LOOCV): tumor versus normal in the peripheral zone ($n = 81$) and the transition zone ($n = 68$), fit separately because normal peripheral- and transition-zone tissue differ systematically in baseline diffusion whereas tumor tissue is comparatively similar across zones, so a pooled boundary would conflate zonal and pathological variation.
Five feature representations were scored through one common pipeline---ADC alone, each individual spectral fraction alone, the two outer fractions $\{D = 0.25,\,3.0\;\umsmm\}$ jointly, the six inner fractions (the two outer compartments removed), and all eight fractions (MAP or NUTS posterior means)---so that single- and multi-feature models are directly comparable.
Within each fold, features were standardized to zero mean and unit variance using statistics from the training ROIs only. Because the eight fractions differ several-fold in scale and cross-ROI variance, this keeps the $L_2$ penalty from shrinking coefficients by feature scale rather than discriminative value and makes the per-bin coefficients (Figure~\ref{fig:sensitivity}) comparable as the change in tumor log-odds per standard deviation of each fraction, a scale-free measure of importance.
The classifier ($C = 1.0$) was then fit on the training ROIs and applied to the held-out ROI, with fitting confined to the training fold to prevent leakage; the conventional in-sample ADC raw-rank AUC is additionally reported as a reference baseline (Table~\ref{tab:auc}).
Performance was the area under the receiver-operating-characteristic curve (AUC), with $95\%$ confidence intervals from patient-level percentile bootstrap (2000 resamples) \citep{efron1993bootstrap}; sensitivity analyses at $C = 0.1$ and $C = 10$ confirmed robustness to regularization strength.

\subsection{ADC Sensitivity Analysis}

The ADC sensitivity vector (Eq.~\ref{eq:sensitivity}) was computed numerically at the average tumor and average normal spectra for each zone, using finite differences with step size $\epsilon = 10^{-4}$.
For the sensitivity-alignment comparison, the logistic-regression \emph{coefficient vector} was taken from a single full-data fit of the eight-feature model---distinct from the leave-one-out detection pipeline above, which evaluates predictive AUC---since only the coefficient direction, not predictive performance, enters this analysis.
The Pearson and Spearman rank correlations between the sensitivity vector and the LR coefficient vector were computed to quantify alignment.
Bootstrap 95\% confidence intervals for ROI-level correlations between ADC and the spectral discriminant were obtained by resampling ROI pairs with replacement (1000 resamples).

% ======================================================================
\subsection{Pixel-wise and Direction-wise Spectral Estimation}

To demonstrate the framework at the voxel level, pixel-wise decomposition was performed on a single axial slice from one patient\stephan{please confirm whether this patient is inside the 149-ROI cohort and that the segmented tumor is peripheral-zone; if the patient is in-cohort, the ``independent/separate dataset'' framing must be dropped, since the classifier is trained on that cohort.}, with the prostate manually segmented. Each voxel's decay, normalized by its $b = 0$ signal, was decomposed by MAP (Eq.~\ref{eq:map}) and NUTS (Eqs.~\ref{eq:prior_R}--\ref{eq:likelihood}) with the ROI-level parameters, and a spectral discriminant score was computed by applying the peripheral-zone classifier coefficients to the pixel spectra (the segmented tumor lay in the peripheral zone); for NUTS, discriminant uncertainty was propagated by applying the classifier to each of 500 posterior draws per voxel, yielding a per-voxel distribution of scores.

To verify the trace-averaging assumption used at the ROI level, a separate acquisition that retained the per-encoding-direction signal---13 ROIs across 7 patients, with the same 15 b-values along three orthogonal directions---was decomposed independently for each direction with the MAP estimator and compared to the trace-averaged spectrum. The aggregate per-direction coefficient-of-variation statistics reported in Results are computed across all 13 ROIs; Figure~\ref{fig:directions} shows a representative subset of four patients (one peripheral-zone and one transition-zone ROI each).

