\section{Results}

\subsection{Diffusivity spectra and identifiability}

The signal-to-noise ratio of the trace decays, computed as $S_0/\sigma$ from the per-ROI NUTS posterior, had a median of 303 across the 149~ROIs (interquartile range 176--478; full range 25--1548).
The wide spread reflects variation in tissue composition and distance from the endorectal coil.
Figure~\ref{fig:spectra} shows the cohort-averaged MAP and NUTS spectral fractions for tumor and normal tissue in both anatomical zones.
Tumor ROIs exhibited an elevated restricted-diffusion water fraction ($D = 0.25\;\umsmm$: 18.5\% MAP / 16.6\% NUTS in PZ tumor vs.\ 7.2\% / 5.7\% in PZ normal) and a reduced free-water fraction ($D = 3.0\;\umsmm$: 18.9\% MAP / 21.0\% NUTS in PZ tumor vs.\ 40.3\% / 47.5\% in PZ normal), consistent with increased cellular density restricting water mobility.
These patterns were qualitatively similar in the transition zone: TZ tumor showed an elevated restricted fraction ($D = 0.25\;\umsmm$: 26.4\% MAP / 25.0\% NUTS vs.\ 9.9\% / 8.5\% in TZ normal) and a reduced free-water fraction ($D = 3.0\;\umsmm$: 12.7\% / 14.5\% vs.\ 29.3\% / 29.9\%). At the tuned regularization, MAP and NUTS agree closely on the two well-identified outer compartments.

We summarize per-component identifiability by the cohort-averaged posterior coefficient of variation (CV; per-ROI values for representative cases in the supplementary atlas, Figure~\ref{fig:atlas}).
Only restricted diffusion ($D = 0.25\;\umsmm$, mean CV = 0.20) and free water ($D = 3.0\;\umsmm$, mean CV = 0.32) had cohort-averaged CV below 0.40.
The IVIM-perfusion proxy ($D = 20.0\;\umsmm$) was moderately constrained (CV = 0.36).
Intermediate components ($D = 0.50$--$1.00\;\umsmm$) had CV $> 0.80$, indicating posterior uncertainty comparable to the estimate itself.
This pattern is grounded in the Fisher information of the acquisition (Figure~\ref{fig:fisher}): the intermediate components are mutually correlated above $0.99$ and are therefore intrinsically ill-conditioned.
Simulation confirms that this is a data limit rather than an inference artifact: on simulated ground truths the sampler recovers smooth spectra accurately at the cohort SNR but only partially recovers a $\delta$-spike placed in the intermediate bins---smearing it into its correlated neighbors---while jointly recovering the noise level $\sigma$ to within $\sim$10\% except on that unrepresentable $\delta$ (Figure~\ref{fig:validation}).

\subsection{Tumor detection}

Table~\ref{tab:auc} and Figure~\ref{fig:roc} summarize detection performance.
ADC was the strongest single detector (raw-rank AUC PZ $0.951$, TZ $0.979$), and the spectral classifiers matched it: through the identical leave-one-out pipeline the eight-fraction logistic regression reached PZ $0.926$ / TZ $0.923$ (NUTS) and PZ $0.917$ / TZ $0.952$ (MAP), with bootstrap confidence intervals overlapping ADC's throughout.
The fair head-to-head is ADC versus the spectral models scored through the same held-out pipeline; the in-sample ADC raw rank is reported as the hardest reference baseline.
Restricting the classifier to the two outer fractions $\{D = 0.25,\,3.0\;\umsmm\}$ did not reduce performance (NUTS PZ $0.933$, TZ $0.937$; MAP TZ $0.965$). The mirror ablation pins down where the signal lives: classifying on the six inner fractions alone---the two outer compartments removed---dropped NUTS detection to PZ $0.817$ / TZ $0.775$ (Table~\ref{tab:auc}). Detection is thus both carried and bounded by the two outer compartments; the inner fractions add only finite-sample noise once those are present, which is why the two-feature model matches, and sometimes slightly exceeds, the eight-feature one.
This reflects redundancy rather than uselessness. The individual-fraction AUCs (Table~\ref{tab:auc}) show the two outer bins are the strongest single detectors (AUC $0.88$--$0.92$), but several intermediate fractions also discriminate on their own (up to $\sim$0.85, varying by zone), and even the four most poorly identified bins ($D = 0.5$--$1.5\;\umsmm$, posterior CV $> 0.7$) reach AUC $0.82$ (PZ) / $0.81$ (TZ) when used together. The spectrum is therefore not split into informative and uninformative bins: the inner fractions co-vary with the dominant tumor axis, so each is individually informative yet contributes nothing once the two outer compartments are present.
Grade-classification AUCs are not reported ($n = 29$ graded tumors are too imprecise; see Table~\ref{tab:auc}); grading is instead presented below as grade-dependent spectral shifts.
Detection AUCs varied by $< 0.03$ across $C = 0.1$--$10$.

\subsection{Why ADC works}

Across ROIs, ADC was near-perfectly anti-correlated with the spectral discriminant---the trained logistic-regression score applied to each ROI's spectrum---in both zones and for both spectral estimators (Figure~\ref{fig:adc_discriminant}). Ranking ROIs by ADC is therefore almost identical to ranking them by the optimal spectral classifier: using NUTS posterior-mean features at $C = 1.0$, the Spearman rank correlation was $\rho = -0.979$ (PZ, $n = 81$) and $\rho = -0.977$ (TZ, $n = 68$), with Pearson correlations of $-0.977$ ($95\%$ bootstrap CI $[-0.985,\,-0.964]$) and $-0.981$ ($[-0.988,\,-0.971]$). Constrained-MAP features gave equivalent values (Spearman PZ $-0.980$, TZ $-0.929$; Pearson $-0.972$, $-0.967$). The near-identity was robust across regularization strengths ($C = 0.1$ to $10$) and held for both MAP and NUTS, indicating that it is a property of the data rather than of any particular classifier setting.

The mechanism is visible in the per-bin structure of the discriminant (Figure~\ref{fig:sensitivity}).
The ADC sensitivity vector $\partial\text{ADC}/\partial R_j$ at the average tumor spectrum is strongly negative for restricted diffusion ($-1.59$: a larger restricted fraction lowers ADC) and positive for free water ($+0.85$); because the logistic-regression coefficients point the opposite way, Figure~\ref{fig:sensitivity} plots the sign-flipped $-\partial\text{ADC}/\partial R_j$ alongside them, so aligned bars indicate matching direction.
Using NUTS features, the two vectors align well (Pearson $r = -0.79$, PZ, $95\%$ CI $[-0.89,\,-0.50]$; $r = -0.88$, TZ, $[-0.95,\,-0.70]$), and the alignment is carried by the two outer bins: only $D = 0.25$ and $D = 3.0\;\umsmm$ have logistic coefficients whose bootstrap intervals exclude zero.
This is not because ADC ignores the intermediate compartments: its sensitivity $-\partial\text{ADC}/\partial R_j$ is appreciable there (e.g.\ $D = 0.5\;\umsmm$), since ADC is a smooth, monotonic-in-$D$ weighting of the entire spectrum. The resolution is that discriminative power is sensitivity \emph{times} tumor--normal contrast: the intermediate fractions carry little contrast beyond what the outer bins already capture (Table~\ref{tab:auc}) and are poorly identified (high CV), so ADC's sensitivity to them is spent on a low-contrast, noisy direction that shifts neither score's ranking. Both summaries therefore reduce to the same two outer compartments, which is why the per-ROI scores rank almost identically (Figure~\ref{fig:adc_discriminant}) even though the per-bin profiles diverge in the middle. The divergence is confined to bins that are diagnostically inert for both---it is not evidence that ADC and the spectral classifier estimate different quantities.
\subsection{Spectral Decomposition by Gleason Grade Group}

Figure~\ref{fig:spectrum_ggg} shows the cohort's average NUTS posterior-mean spectrum by Gleason Grade Group, with normal tissue ($n = 109$) as the shared baseline and tumor resolved into GGG\,$=\,1$ ($n = 8$), GGG\,$=\,2$ ($n = 12$), and GGG\,$\geq 3$ ($n = 9$).
\stephan{exploratory subsection --- please confirm the intermediate-bin interpretation below; the independent grading-axis claim is dropped for now.} These grade-stratified spectra are exploratory: the small, unevenly graded subgroups (down to $n = 8$) do not support a grading axis independent of detection, but the per-bin trends below are suggestive.
The free-diffusion bin at $D = 3.0\;\umsmm$ (intact gland lumen) is suppressed as soon as tumor is present ($0.39$ normal $\to 0.21$ at GGG\,$=\,1$) and then changes little ($0.16$ at GGG\,$\geq 3$): it saturates at tumor onset, separating tumor from normal but not grade from grade.
By contrast the restricted-cellular bin ($D = 0.25\;\umsmm$) rises with grade ($0.07$ normal to $0.31$ at GGG\,$\geq 3$) and the glandular-epithelial bin ($D = 2.0\;\umsmm$) is preserved in low-grade tumor then depleted ($0.26$ normal, $0.28$ at GGG\,$=\,1$, $0.17$ at GGG\,$\geq 3$): these bins, where ADC is least sensitive (Figure~\ref{fig:sensitivity}), track aggressiveness.
Grade-specific sample sizes are small (down to $n = 8$--$9$) and the corresponding confidence bands wide, and 11 of the 40 tumor ROIs (7 with biopsy Gleason Grade Group 0 and 4 ungraded) lack a Grade Group and are excluded from this grade stratification, though they remain tumor ROIs in the detection analysis.
Method robustness: re-running the same group averaging on tuned-MAP ($\lambda = 10^{-3}$) posterior estimates rather than NUTS posterior means yielded indistinguishable group trajectories on all eight bins.
% ======================================================================
\subsection{Prediction confidence}

Each classifier returns a probability $P(\mathrm{tumor})$ rather than a hard label, and misclassified ROIs concentrate near the decision boundary ($P \approx 0.5$): ranking predictions by their distance from the boundary separates likely errors from confident, correct calls, so the classifier flags its own least-reliable predictions (Figure~\ref{fig:uncertainty}).
This confidence is a property of the logistic model, available from the point estimate alone, and is shared by the ADC- and spectrum-based classifiers.
Propagating the full NUTS posterior through the classifier---applying it to every posterior draw rather than the posterior mean---does not add to this.
The 90\% credible interval on $P(\mathrm{tumor})$ is on average $2.4\times$ wider for misclassified than for correctly classified ROIs (pooled across zones, mean width $0.39$ vs $0.16$), but this reflects the logistic geometry rather than the spectral posterior: the slope $P(1-P)$ is largest at the boundary, so in logit space the difference falls to $\approx 1.3\times$ and is no longer significant, and the interval width does not predict misclassification once distance to the boundary is accounted for.
Offering the per-component posterior standard deviations to the classifier as features likewise leaves detection AUC essentially unchanged.
The posterior therefore quantifies which spectral components are reliably estimated (identifiability), but adds no diagnostic information to the tumor-versus-normal decision beyond the point estimate.

% ======================================================================
\subsection{Pixel-wise Spectral Decomposition}

Pixel-wise spectral estimation was performed on the manually segmented prostate of the supplementary patient (MAP and NUTS; all NUTS voxels converged, $\hat{R} < 1.05$).
At the voxel level the restricted-diffusion ($D = 0.25\;\umsmm$) and free-water ($D = 3.0\;\umsmm$) fraction maps were spatially anti-correlated (Figure~\ref{fig:pixelwise}), and the per-voxel spectral discriminant (ROI-level PZ coefficients applied to the pixel spectra) reproduced the ADC pattern (anti-correlation $r = -0.97$).
The contribution unique to the Bayesian decomposition is the accompanying per-voxel uncertainty map: propagating the NUTS posterior through the classifier flags where a voxel's score is least reliable, and the eight-bin detector is on average $1.8\times$ more uncertain than the two-bin detector because it weights the unidentifiable intermediate compartments---an uncertainty that ADC and point-estimate methods cannot provide.
As at the ROI level, the discriminant adds no detection signal beyond ADC; its value here is the calibrated, spatially-resolved confidence it carries.
Direction independence was verified at the whole-ROI level on a separate three-direction acquisition (Figure~\ref{fig:directions}): across all 13 regions the per-direction coefficient of variation of the estimated spectrum was lowest for the diagnostically dominant restricted-diffusion bin ($D = 0.25\;\umsmm$, mean $9\%$) and was large only on the near-empty bins ($D = 20\;\umsmm$), where it reflects a small-denominator effect rather than genuine anisotropy, confirming that trace averaging is appropriate for spectral decomposition.
